SummaryStats

### 2023-03-01 analysis/code/Exploration.Rmd
#1 update exploration + housekeeping

There are no duplicate rows, but there are 3708 observations with duplicate city-year keys. These observations make up 0.3% of the data set. I can probably drop them? Unless they are very different from the rest for some reason. *0.337% is suspiciously close to the share of city-years with extinctions.*

```{r get obs with dupe keys}

# if there are multiple observations for a city-year key, number them
data_count_dupes <- data %>% 
  group_by(city_id, year) %>% 
  mutate(count_dupe = row_number()) %>% 
  ungroup()

# show that all observations with duplicate keys have terr_id A1341
data_count_dupes %>% filter(count_dupe != 1) %>% ggplot() + geom_bar(aes(x=terr_id))

# show that exactly half of all A1341 observations are dupes
data_count_dupes %>% 
  filter(terr_id == "A1341") %>% 
  mutate(count_dupe = factor(count_dupe)) %>% 
  ggplot(aes(x=count_dupe)) + 
  geom_bar() +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=0)

# suspicion: the only thing that varies for dupes is `X` (auto-generated while joining)
# and `count_diff` (only calculated for the first instance of a city-year observation).
data %>% filter(terr_id == "A1341") %>% select(-X, -count_diff) %>% distinct() %>% count()

Conclusion: for some reason, all observations with terr_id "A1341" got duplicated somewhere along the way. I *could* just drop all the dupes, because the `count_diff` is always with the first version:

```{r}
data_count_dupes %>% 
  filter(terr_id == "A1341") %>% 
  mutate(count_dupe = factor(count_dupe)) %>% 
  ggplot(aes(x=count_dupe)) + 
  geom_boxplot(aes(y=count_diff))
```

But because this isn't actually the final version of the data, and this might happen again, it's probably prudent to check where the duplication occurs. I do this in a separate Rmd.

@@

```{r lineage level data}
terrs <- data %>% 
  distinct(terr_id, year, .keep_all=TRUE) %>% 
  select(terr_id, year, territory, count_cities, final_year, extinction)
```

Note: this only keeps the first city-year observation for each lineage-year.
However, we only use this data set to look at variables that are identical for 
all city-years in a lineage-year, so `count_cities` and `extinction`.

@@

Left: Is the lifespan of a lineage correlated to its (maximum lifetime) size? 
Right: Are lineages less likely to go extinct when they are larger? (extinction on contemp. size)

```{r}
lifespan_on_max_size <- terrs_life %>% 
  ggplot(aes(x=log(max_size), y=lifespan))+
  geom_point()+
  geom_smooth(method="lm", se=FALSE)

@@

#mod_lifespan_size <- lm(lifespan ~ log(max_size), data=terrs_life)
#summary(mod_lifespan_size)
```

### Example cities

>>>>

Yes. each observation has a unique city-year key.

@@

```{r lineage level data}
terrs <- data %>% 
  distinct(terr_id, year, .keep_all=TRUE) %>% 
  select(terr_id, year, territory, count_cities, final_year, extinction)

# terrs <- read_csv("build/output/lineages.csv") # comment out for sanity check
```

Note: this only keeps the first city-year observation for each lineage-year.
However, we only use this data set to look at variables that are identical for 
all city-years in a lineage-year, so `count_cities` and `extinction`. I did a sanity check: the following plots look exactly the same for the lineages.csv data.

@@

#### Correlation between size and survival

```{r}
#mod_lifespan_size <- lm(lifespan ~ log(max_size), data=terrs_life)
#summary(mod_lifespan_size)

lifespan_on_max_size <- terrs_life %>% 
  ggplot(aes(x=log(max_size), y=lifespan))+
  geom_point()+
  geom_smooth(method="lm", se=FALSE)

@@

Left: The lifespan of a lineage is weakly correlated to its maximum lifetime size.
Right: This honestly doesn't say much. What it tries to say is that lineages may be less likely to go extinct when they are larger.

### Cities

#### Lifetime summary statistics for cities

```{r}
cities_life <- data %>% 
  group_by(city_id) %>% 
  summarise(
    years_exists = max(year) - min(year),
    rulers_count = n_distinct(terr_id),
    extinctions_count = sum(extinction),
    diffs_count = sum(count_diff != 0, na.rm=TRUE),
    mean_integ = mean(count_cities),
    lifetime_integ_diff = sum(count_diff, na.rm=TRUE),
    mean_yearly_diff = mean(count_diff, na.rm=TRUE),
    sd_yearly_diff = sd(count_diff, na.rm=TRUE)
    )

# check whether diffs_count works:
# only1001 <- data[data$city_id == 1001,]
# sum(only1001$count_diff != 0, na.rm = TRUE) # there are 57 nonzero countdiffs
# cities_life %>% filter(city_id == 1001) %>% select(diffs_count) # yes it works

summary(cities_life)
```

This is very insightful. Some takeaways:

- I have almost 500 observations on average for each city.
- **A normal city changes hands 3-4 times throughout its history.**
- **Of these territory changes, 1-2 are due to extinction.**
- **A normal city undergoes around 60 changes in integration.**
- Less than a quarter of cities is integrated with less than 100 other cities on average throughout its history.
- Most cities become drastically more integrated over time.
- Yearly differences in integration are positive on average but vary a lot.
